Synthesizing Quotient Lenses

### 2018-03-05 papers/QuotientLenses/main.tex
pass up to 3.1

\lstset{
language=Caml, 
basicstyle=\upshape\sffamily,
keywordstyle=\upshape\sffamily\color{dkblue},
keepspaces=true,
framexleftmargin=1ex,
framexrightmargin=1ex,
showstringspaces=false,
commentstyle=\itshape\rmfamily,
emph={synth,project,perm,squash,normalize,using,ins,del,lens},
emphstyle=\upshape\sffamily\color{dkblue},
columns=flexible,
% BCP: I find this distracting:
% stringstyle=\sffamily\color{dkred},
mathescape,
 }

@@

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\kw}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\codefont}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\project}[2]{\ensuremath{\kw{project} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\kw{squash} \; #1 \rightarrow #2
\kw{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\kw{perm}\; (#1)\; \kw{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\kw{normalize} \; (#1, #2, #3)}}
\newcommand{\sep}{\ensuremath{\ | \ }}
\newcommand{\canonizer}{\ensuremath{\kw{Canonizer}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\kw{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\kw{put}}}
\newcommand{\create}{\ensuremath{\kw{create}}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}

@@

Oftentimes, the same data is stored in a variety different formats. For
example, \bibtex{} and EndNote both store bibliography data; spreadsheet data
can be represented as a CSV file or a TSV file; web APIs can return data as a
JSON or
XML object. When different applications manipulate the same data in different formats,
synchronizing the applications amounts to synchronizing their respective data
formats. One way to synchronize data formats is to use \emph{lenses}, \emph{i.e.,} programs that represent pairs of functions for translating data from
format A to B (the \emph{get} direction) and back again (the \emph{put}
direction).  These lenses, when defined using domain-specific
languages like Boomerang~\cite{boomerang} or
BiGul~\cite{ko2016bigul}, typically come with a variety of guarantees,
the \emph{lens laws}, that help ensure put and get functions are effective
inverses of one another.
%comprise such synchronization functions, and the ``lens
%laws'' provide restrictions on how these synchronization functions interoperate%.
%Bidirectional languages like Boomerang~\cite{boomerang} and
%BiGul~\cite{DBLP:conf/pepm/KoZH16} use a single expression to encode all of the
%lens's synchronization functions at the same time.

One way to classify lenses is by the laws they obey.  Amongst the most
restrictive lenses are the \emph{bijective lenses}.  These lenses demand
that get and put functions form a bijection, meaning no information
is lost when one translates from one data format to another.  While this
strong guarantee can be useful in some contexts, it is more often the case
that some of the information in one or both of the data formats is superfluous.
For instance, one typically does not care to preserve the amount of
whitespace between characters.  In other circumstances, capitalization
or the order of fields in a record or items in a list may be unimportant.
%
%bijective lenses have
%strong correctness guarantees, they are often
%However, these lens laws are too restrictive for many important transformations.
%The bijective lens laws do not permit transformations that ignore this
%unimportant information---.
Many data formats have other ad hoc rules as well.
Returning to our bibliography example, recall that in \bibtex{},
an author can be specified as
``\cd{lastname, firstname}'' or as ``\cd{firstname lastname}''---it does not matter which of the
two formats is used.
%While it is important to allow authors
%to be input in both ways, we would like these authors to only be output as
%\cd{firstname lastname}, but the lens laws prevent such functions.
%
{\em Quotient lenses}~\cite{quotientlenses} loosen many of the restrictions
of bijective lenses by introducing an equivalence relation and considering
bijections modulo that relation.  Such quotient lenses make it possible
to align many more data sources than would otherwise be possible by
ignoring details that a programmer views as inessential.
%each of the data formats is quotiented by some equivalence relation and the lens
%laws are taken modulo equivalence.
%This allows lenses to be simpler and more
%flexible. For example, inessential details like whitespace and orderings of
%fields can be ignored when writing the transformations between the formats.

%However, while plain lenses are widely used in both standalone systems and
%popular libraries, Boomerang~\cite{boomerang} is the only implementation
%currently offering quotient lenses. Even though quotient lenses permit simpler
%transformations between the formats, they merely push complexity elsewhere.
Ordinary bijective lenses are fairly tricky to program, especially for
beginners, as they require programmers to think in two directions at once
and learn a new type system, which comes with some tricky ambiguity constraints.
Unfortunately, quotient lenses do not improve the situation:  To
write a quotient lens in Boomerang, programmers must work with standard lens
combinators as well as a system of \emph{canonizers}.
These canonizers encode two functions, one that
maps data represented in the original external format
to a canonical element, and another, in the opposite direction, that chooses
which elements of the original format correspond to values in the canonical
format. Furthermore, if a user wants to define the domain and range of
a lens explicitly (a programming practice that is encouraged), she must
provide regular expressions defining input and output formats.
Consequently, to define a quotient lens, one must often
construct three different sorts
of terms: (1) regular expressions that define the two data formats
to be processed, (2) canonizers to normalize the data
and choose representatives, and (3) a bijective lens that maps
one canonical form into another.  Somewhat frustratingly,
these three terms share a fair amount of structure.
%Moreover, writing the lens between the equivalence classes can be
%challenging due to the rather fiddly typechecking, mainly due to
%``non-ambiguity'' conditions in the lens well-formedness rules.

@@

Second, we use QREs to define a class of quotient lenses called {\em QRE
lenses}, and, we show how to synthesize such QRE
lenses automatically.
More specifically, in
order to generate a lens, a programmer provides a pair of QREs
describing the source and target formats, and, optionally,
a suite of examples that demonstrates how the
lens should act on specific examples.  Given this information,
we are able to exploit and extend recent work on algorithms for synthesis of
bijective lenses~\cite{optician} to generate quotient lenses.
Importantly, the lenses generated by
our synthesis algorithm are guaranteed to translate data back and forth
between source and target QREs and to satisfy the provided examples.
Moreover, while our synthesized quotient lenses have a relatively
restricted form (a fact that cuts down the amount of non-determinism in
a specification and thus helps speed up synthesis), we prove that
this rigidity does not compromise expressiveness.

The following list summarizes the central contributions of our work.
\begin{enumerate}
  \item We introduce {\em Quotient Regular Expressions} (QREs),
  a compact, convenient notation for defining (1) a
  regular language modulo some equivalence relation, and (2) a canonizer
  for that relation (Section~\ref{QRE}).
  \item We define a language of {\em QRE lenses}.  This language is
  a subset of the full set of bijective quotient
  lenses whose types are given by QREs. Our main technical contribution
  is a proof that our QREs are just as expressive as the full set of bijective
  quotient lenses.
  % In other words, if there eeach QRE lens $q$ defines the same transformation as a
  %QRE lens $q'$ that is in normal form.
  (Section~\ref{QRE-lenses}).
  \item We reduce the problem of {\em synthesizing}
  QRE lenses to the previously studied problem of synthesizing bijective lenses
  between regular languages (Section~\ref{synth}).
  \item We extend the Boomerang {\em implementation} with QREs
  and QRE lens synthesis.  We demonstrate the utility of our
  implementation by using it to
  synthesize QRE lenses between data formats drawn from the
  {\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} discusses related work and~\ref{concl} summarizes the
results of the paper.

@@

\emph{Lenses} define computations that can be read in two directions,
the get direction (forwards) and the put direction (backwards).  The simplest lenses define bijections between source and
data, but we are interested in \emph{quotient lenses}, which define
bijections modulo some equivalence relation.  The remains of this section
describe and motivate a new language and system for defining such lenses.

@@

One natural way to define lenses that will synchronize data in these formats
is to provide the following items: \dpw{this is slightly redundant from the intro.  It
adds a little more info, but not much.  Should this be cut?  Should the intro be expanded
slightly?  Should it be left as is (more or less)?}

\begin{enumerate}
  \item A pair of descriptions of the external source and target formats.
  Such descriptions may be given by regular expressions, for instance.
  They serve as documentation to describe the domains over which the lens
  is to operate.
  \item A pair of \emph{canonizers} that map external source (and target)
  objects into internal, canonical forms.  Such canonizers will take
  care of stripping away irrelevant information and otherwise normalizing
  differences between equivalent representations of the same logical data.
  When given an internal representation, such canonizers must also operate
  in the reverse direction, \emph{choosing} an element in the external format
  to represent the canonical internal element.
  \item A pair of descriptions of the internal source and
  target formats, again, to document the structure of the transformation.
  \item A bijective lens to map back and forth between canonical elements
  of the internal source and target formats.
\end{enumerate}

\noindent
Indeed, Boomerang's quotient lenses are structured this
way~\cite{quotientlenses}.
Of course, some of these elements are optional.  For instance, it is
not necessary to document internal and external formats.  However, we do
view it as good engineering practice to do so, as such documentation makes
lenses easier to read, understand, use and maintain.  In this case, there is
significant redundancy here:  External and internal formats
typically have some overlap; external formats and canonizers have related
structure; and the structure of bijective lens computations is largely driven
by the structure of the internal formats.  Hence, overall, a more compact
notation would simplify description and legibility of such computations.
In addition,
the lens and canonizer languages themselves can be somewhat fiddly and
difficult to
master.  For instance, to quotient out whitespace inside, and in between,
fields while also ignoring field order, one must nest canonizing combinators
within one another in a somewhat tricky way.

@@

A {\em Quotient Regular Expression} (QRE) defines an external regular
expression $R$, an equivalence relation $\sim$ on the language
described by $R$, and a canonical set of representatives of $\sim$,
all at once.  QREs are constructed from a simple set of primitives
and composed using familiar regular operators
(choice `\cd{|}', concatenation `\cd{.}',
and Kleene star `$^*$').  Users may also treat QREs as pairs of
get and put functions and compose such pairs (the functional
composition operator `;').

For example, to begin construction of our \bibtex{} format, we might
first define a whitespace format, which, externally, matches
any non-zero number of whitespace characters, and converts
any such whitespace into a single space character (its canonical form).
Such a definition uses the \kw{project} primitive.

@@

Sometimes there are multiple disjoint representations of the same data.
In such situations, the \kw{squash} combinator creates a QRE that
allows external data to be in either format, and converts any
data in one format to the other.
For instance, assume
the \codefont{comma\_name} format describes ``Doe, John''
and the \codefont{space\_name} format describes ``John Doe''
and \codefont{s\_to\_c} is a function from the first to the second.  In this case,
the following instance of \kw{squash} creates the desired canonizer.

@@

One way to define the function used by squash is simply to write it
from scratch in OCaml.  However, our system will also synthesize
such functions automatically from pairs of regular languages and
examples.
%
\begin{lstlisting}
let l = synth comma_name $\Leftrightarrow$ space_name using {("Doe, John", "John Doe")}
let s_to_c = get l
\end{lstlisting}
%
The first line above synthesizes a \emph{bidirectional} lens between
\codefont{comma\_name} and \codefont{space\_name} using the listed example
transformation as a guide.  The second line simply
extracts the get direction transformation from the lens,
which is what we need for squash.

Another QRE primitive is the composition combinator ``;''. This combinator
applies two equivalence relations to a regular language in sequence. For
example, suppose that we wish to express a name as $\kw{last\_name \cdot
``,\text{''} \cdot \text{wsp+} \cdot first\_name}$, or as $\kw{first\_name\cdot
\text{wsp+} \cdot last\_name}$, with $\kw{first\_name\cdot \text{wsp+} \cdot
last\_name}$ as the canonical representation. We also wish to project away the
extra whitespace in between \kw{first\_name} and \kw{last\_name} once this
identification has been made. Then we can express this compound equivalence
relation by a composition of a \textit{squash} QRE and a QRE that projects away
the extra whitespace in between \kw{first\_name} and \kw{last\_name}

A third useful QRE primitive is the permutation combinator.  For example,
the following instance of \kw{perm} allows label, author and title fields
(which we assume have been defined earlier) to appear in any order.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title)
\end{lstlisting}
%
To normalize any repeated separator between fields, one can extend a permutation
with any other QRE.  For instance, below, we normalize whitespace between fields,
leaving only a single newline.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title) with (project (wsp+ . "," . wsp+) $\to$ ",\n")
\end{lstlisting}
%
The final QRE combinator is the normalize combinator.
This combinator allows a programmer to manually define a function
$f$ which sends each string that matches a regular expression $R$ to some
canonical representative in another regular expression $R'$. The equivalence
relation defined by the normalize combinator is hence the equivalence relation
defined by the {\em fibres} of $f$; that is, for all strings $s$ and $s'$ that
match $R$, $s$ is equivalent to $s'$ if and only if $f(s) = f(s')$.

For instance, assume that $f(s)=$ `` '' for all whitespace strings $s$. Then the
project QRE \kw{p_wsp} which matches any non-zero number of whitespace
characters and converts any such whitespace into a single space character can
be expressed using the normalize combinator as
\begin{lstlisting}
let p_wsp = normalize (wsp+, " ", f)
\end{lstlisting}
The QRE for the bibtex format is presented in its entirety in
Figure~\ref{fig:example-qre}.
\begin{figure}[t]
\begin{lstlisting}
let wsp = [ \n\t\r]

@@

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_fields = perm (label, bib_author, bib_title) with (project (wsp+ . "," . wsp+) $\to$ ",\n")
let bibtex = "@book{" . bib_fields . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
\label{fig:example-qre}
\end{figure}
\subsection{QRE Lenses}
QREs allow users to specify regular languages modulo equivalence relations,
a first step in writing many data transformations.  However, writing the
inner bijective lens itself still takes a fair amount of work.  Each such lens
is constructed from simple primitive combinators including:
%
\begin{itemize}
  \item \kw{del} $s$: delete the constant string $s$ in the get direction;
  produce the constant string $s$ in the put direction.
  \item \kw{ins} $s$: produce the string $s$ in the get direction;
  delete the string $s$ in the put direction.
  \item \kw{copy} $R$: copy the text matching the regular expression $R$ in both
  directions.
\end{itemize}
In addition, such primitives may be combined using the regular operators
Kleene star, alternation and concatenation, or functional composition of
lenses.  Figure~\ref{fig:example-lens} illustrates the use of these
combinators to transform \bibtex{} to EndNote (and vice versa).
A fair amount of trickiness in writing these transformations as users
must keep mental track of what they have parsed so far \emph{in both
directions}, carefully inserting and deleting the relevant syntax on both
sides.

@@

%% The lens shown in Figure~\ref{fig:example-lens} transforms \bibtex{} data to
%% EndNote data  (and vice versa) by matching the delimiters and separators in a
%% bijective manner. For example, the \cd{preamble} lens replaces the \bibtex{}
%% preamble \verb|@Book| with the EndNote preamble \verb|%0 Book\n%F|. When the
%% programmer is writing the lens, she has to make sure that all of these small
%% deletions and insertions are correct, which can be a challenging task if the
%% lens is large enough, as in the definition of the lens \verb!author_lens! in
%% Figure~\ref{fig:example-lens}. This makes lens programming tedious and
%% error-prone.
Consequently, we offer the programmer some relief by identifying a class of
quotient lenses that we call {\em QRE lenses}, and by showing how to {\em synthesize}
QRE lenses from a pair of QREs and a set of input-output example pairs. Our key
insight is to reduce the problem of synthesizing QRE lenses to the problem of
synthesizing {\em bijective lenses}~\cite{optician}, a simpler subset of
lenses that support strict bijections, rather than bijections modulo
equivalence relations.

@@

This technical investigation justifies replacing manually constructed lenses that may use
functional composition and nested canonizers with synthesized lenses in normal form.
The end result is that users may synthesize complex transformations with a single line
of code.  For instance, all the code in Figure~\ref{fig:example-lens} may be replaced by
a single call to the synthesis prodedure:
\begin{lstlisting}
let bib_to_end : (lens in bibtex $\Leftrightarrow$ endnote)
= synth bibtex $\Leftrightarrow$ endnote  using  {(bib_example, end_example)}
\end{lstlisting}
\noindent Here, the generated  quotient lens synchronizes \cd{bibtex} and
\cd{endnote} formats.  It uses \cd{bib_example} and \cd{end_example} (the two
concrete example strings givn at the beginning of this section) to disambiguate.
In addition. as we saw earlier with the definition of \cd{s_to_c}, the synthesis
procedure itself can be used to create lenses that are in turn used to define
other QREs. The ability to mix QRE specification with QRE lens synthesis
yields a powerful and flexible way of creating bidirectional transformations.

>>>>

\lstset{
language=Caml, 
basicstyle=\linespread{0.8}\upshape\sffamily,
keywordstyle=\upshape\sffamily\color{dkblue},
keepspaces=true,
framexleftmargin=1ex,
framexrightmargin=1ex,
showstringspaces=false,
commentstyle=\itshape\rmfamily,
emph={synth,project,perm,squash,normalize,using,ins,del,lens},
emphstyle=\upshape\sffamily\color{dkblue},
columns=flexible,
% BCP: I find this distracting:
% stringstyle=\sffamily\color{dkred},
mathescape,
 }

@@

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\kw}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\codefont}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\project}[2]{\ensuremath{\kw{project} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\kw{squash} \; #1 \rightarrow #2\; \kw{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\kw{perm}\; (#1)\; \kw{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\kw{normalize} \; (#1, #2, #3)}}
\newcommand{\sep}{\ensuremath{\ | \ }}
\newcommand{\canonizer}{\ensuremath{\kw{Canonizer}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\kw{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\kw{put}}}
\newcommand{\create}{\ensuremath{\kw{create}}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}

@@

Oftentimes, the same data is stored in a variety different formats. For
example, \bibtex{} and EndNote both store bibliography data; spreadsheet data
can be represented as a CSV file or a TSV file; web APIs can return data as a
JSON or
XML object. When different applications manipulate the same data in different formats,
synchronizing the applications amounts to synchronizing their respective data
formats. One way to synchronize data formats is to use \emph{lenses}, \emph{i.e.,} programs that represent pairs of functions for translating data from
format A to B (the \emph{get} direction) and back again (the \emph{put}
direction).  These lenses, when defined using domain-specific
languages like Boomerang~\cite{boomerang} or
BiGul~\cite{ko2016bigul}, typically come with a variety of guarantees,
\emph{lens laws}, that help ensure put and get functions are effective
inverses of one another.%comprise such synchronization functions, and the ``lens
%laws'' provide restrictions on how these synchronization functions interoperate%.
%Bidirectional languages like Boomerang~\cite{boomerang} and
%BiGul~\cite{DBLP:conf/pepm/KoZH16} use a single expression to encode all of the
%lens's synchronization functions at the same time.

One way to classify lenses is by the laws they obey.  Amongst the most
restrictive lenses are the \emph{bijective lenses}.  These lenses demand
that get and put functions form a bijection, meaning no information
is lost when one translates from one data format to another.  While this
strong guarantee can be useful in some contexts, it is more often the case
that some of the information in one or both of the data formats is superfluous.
For instance, one typically does not care to preserve the amount of
whitespace between characters.  In other circumstances, capitalization
or the order of fields in a record or items in a list may be unimportant.
%
%bijective lenses have
%strong correctness guarantees, they are often
%However, these lens laws are too restrictive for many important transformations.
%The bijective lens laws do not permit transformations that ignore this
%unimportant information---.
Many data formats have other ad hoc rules as well.
In bibliographies, for example, recall that in \bibtex{},
an author can be specified as
``\cd{lastname, firstname}'' or as ``\cd{firstname lastname}''---it does not matter which of the
two formats is used.
%While it is important to allow authors
%to be input in both ways, we would like these authors to only be output as
%\cd{firstname lastname}, but the lens laws prevent such functions.
%
{\em Quotient lenses}~\cite{quotientlenses} loosen many of the restrictions
of bijective lenses by introducing an equivalence relation and considering
bijections modulo that relation.  Such quotient lenses make it possible
to align many more data sources than would otherwise be possible by
ignoring details that a programmer views as inessential.
%each of the data formats is quotiented by some equivalence relation and the lens
%laws are taken modulo equivalence.
%This allows lenses to be simpler and more
%flexible. For example, inessential details like whitespace and orderings of
%fields can be ignored when writing the transformations between the formats.

%However, while plain lenses are widely used in both standalone systems and
%popular libraries, Boomerang~\cite{boomerang} is the only implementation
%currently offering quotient lenses. Even though quotient lenses permit simpler
%transformations between the formats, they merely push complexity elsewhere.
Ordinary bijective lenses are fairly tricky to program, especially for
beginners, as they require programmers to think in two directions at once
and to learn a new type system, which comes with some tricky ambiguity constraints.
Unfortunately, quotient lenses do not improve the situation:  To
write a quotient lens in Boomerang, programmers must work with standard lens
combinators as well as a system of \emph{canonizers}.
These canonizers encode two functions, one that
maps data represented in the original external format
to a canonical element, and another, in the opposite direction, that chooses
which elements of the original format correspond to values in the canonical
format. Furthermore, if a user wants to define the domain and range of
a lens explicitly (a programming practice that is encouraged), she must
provide regular expressions defining input and output formats.
Consequently, to define a quotient lens, one must often
construct three different sorts
of terms: (1) regular expressions that define the two data formats
to be processed, (2) canonizers to normalize the data
and choose representatives, and (3) a bijective lens that maps
one canonical form into another.  Somewhat frustratingly,
these three terms share a fair amount of structure.
%Moreover, writing the lens between the equivalence classes can be
%challenging due to the rather fiddly typechecking, mainly due to
%``non-ambiguity'' conditions in the lens well-formedness rules.

@@

Second, we use QREs to define a class of quotient lenses called {\em QRE
lenses}, and, we show how to synthesize such QRE
lenses automatically.
More specifically, in
order to generate a lens, a programmer provides a pair of QREs
describing the source and target formats, and, optionally,
a suite of examples that demonstrates how the
lens should act on certain inputs.  Given this information,
we are able to exploit and extend recent work on algorithms for synthesis of
bijective lenses~\cite{optician} to generate quotient lenses.
Importantly, the lenses generated by
our synthesis algorithm are guaranteed to translate data back and forth
between source and target QREs and to satisfy the provided examples.
Moreover, while our synthesized quotient lenses have a relatively
restricted form (a fact that cuts down the amount of non-determinism in
a specification and thus helps speed up synthesis), we prove that
this rigidity does not compromise expressiveness.

The following list summarizes the central contributions of our work.
\begin{enumerate}
  \item We introduce {\em Quotient Regular Expressions} (QREs),
  a compact, convenient notation for simultaneously defining a
  regular language modulo some equivalence relation, and a canonizer
  for that relation (Section~\ref{QRE}).
  \item We define a language of {\em QRE lenses}.  This language is
  a subset of the full set of bijective quotient
  lenses whose types are given by QREs. Our main technical contribution
  is a proof that our QREs are just as expressive as the full set of bijective
  quotient lenses.
  % In other words, if there eeach QRE lens $q$ defines the same transformation as a
  %QRE lens $q'$ that is in normal form.
  (Section~\ref{QRE-lenses}).
  \item We reduce the problem of {\em synthesizing}
  QRE lenses to the previously studied problem of synthesizing bijective lenses
  between regular languages (Section~\ref{synth}).
  \item We extend the Boomerang {\em implementation} with QREs
  and QRE lens synthesis.  We demonstrate the utility of our
  implementation by using it to
  synthesize QRE lenses between data formats drawn from the
  {\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} discusses related work and~\ref{concl} summarizes the
results of the paper.

@@

\emph{Lenses} define computations that can be read in two directions,
the get direction (forwards) and the put direction (backwards).  The simplest lenses define bijections between source and
data, but we are interested in \emph{quotient lenses}, which define
bijections modulo some equivalence relation.  The remainder of this section
describes and motivates a new language and system for defining such lenses.

@@

One natural way to define lenses that will synchronize data in these formats
is to provide the following items: \dpw{this is slightly redundant from the intro.  It
adds a little more info, but not much.  Should this be cut?  Should the intro be expanded
slightly?  Should it be left as is (more or less)?} \saz{I think that this is
not too redundant and that it can be left as-is, modulo space crunching.}

\begin{enumerate}
  \item A pair of descriptions of the external source and target formats, here
    \bibtex{} and EndNote.
  Such descriptions may be given by regular expressions, for instance.
  They serve as documentation to describe the domains over which the lens
  is to operate.
  \item A pair of \emph{canonizers} that map external source (and target)
  objects into internal, canonical forms.  Such canonizers will take
  care of stripping away irrelevant information and otherwise normalizing
  differences between equivalent representations of the same logical data.
  When given an internal representation, such canonizers must also operate
  in the reverse direction, \emph{choosing} an element in the external format
  to represent the canonical internal element.
  \item A pair of descriptions of the internal source and
  target formats, again, to document the structure of the transformation.
  \item A bijective lens to map back and forth between canonical elements
  of the internal source and target formats.
\end{enumerate}

\noindent
Indeed, Boomerang's quotient lenses are structured this
way~\cite{quotientlenses}.
Of course, some of these elements are optional.  For instance, it is
not necessary to document internal and external formats.  However, we do
view it as good engineering practice to do so, as such documentation makes
lenses easier to read, understand, use and maintain.  In this case, there is
significant redundancy here:  External and internal formats
typically have some overlap; external formats and canonizers have related
structure; and the structure of bijective lens computations is largely driven
by the structure of the internal formats.  Hence, overall, a more compact
notation would simplify description and legibility of such computations.
In addition,
the lens and canonizer languages themselves can be somewhat fiddly and
difficult to
master.  For instance, to quotient out whitespace inside, and between,
fields while also ignoring field order, one must nest canonizing combinators
in a somewhat tricky way.

@@

A {\em Quotient Regular Expression} (QRE) defines an external regular
expression $R$, an equivalence relation $\sim$ on the language
described by $R$, and a canonical set of representatives of $\sim$,
all at once.  QREs are constructed from a simple set of primitives
and composed using familiar regular operators
(choice `\cd{|}', concatenation `\cd{.}',
and Kleene star `$^*$').  % Users may also treat QREs as pairs of
% get and put functions and compose such pairs using the functional
% composition operator `;'.

For example, to begin construction of our \bibtex{} format, we might
first define a whitespace format, which, externally, matches
any non-zero number of whitespace characters, and converts
any such whitespace into a single space character (its canonical form).
This whitespace-normalizing QRE is easily defined by using the \kw{project} primitive:

@@

Sometimes there are multiple disjoint representations of the same data.
In such situations, the \kw{squash} combinator creates a QRE that
allows external data to be in either format, and converts any
data in the first format to the second.
For instance, assume that
the \codefont{comma\_name} format describes ``Doe, John''
and that the \codefont{space\_name} format describes ``John Doe''
and \codefont{s\_to\_c} is a function from the first to the second.  In this case,
the following instance of \kw{squash} creates the desired canonizer.

@@

One way to define the \codefont{s\_to\_c} canonization function used by squash
is simply to write it from scratch. (Boomerang canonizers are written in OCaml.)
However, our system will also synthesize such functions automatically from pairs
of regular languages and examples---here, \codefont{s\_to\_c} is the
\codefont{get} direction of a lens that can be synthesized by:
%
\begin{lstlisting}
let l = synth comma_name $\Leftrightarrow$ space_name using {("Doe, John", "John Doe")}
let s_to_c = get l
\end{lstlisting}
%
The first line above synthesizes a \emph{bidirectional} lens between
\codefont{comma\_name} and \codefont{space\_name} using the listed example
transformation as a guide.  The second line simply
extracts the get direction transformation from the lens,
which is what we need for \kw{squash}.

Another QRE primitive is the sequential composition combinator ``;''.  For
example, suppose that we wish to express a name as $\kw{last\_name \cdot
``,\text{''} \cdot \text{wsp+} \cdot first\_name}$, or as $\kw{first\_name\cdot
\text{wsp+} \cdot last\_name}$, with $\kw{first\_name\cdot \text{wsp+} \cdot
last\_name}$ as the canonical representation. We also wish to project away the
extra whitespace in between \kw{first\_name} and \kw{last\_name} once this
identification has been made. Then we can express this compound equivalence
relation by a composition of a \kw{squash} QRE and a QRE that projects away
the extra whitespace in between \codefont{first\_name} and \codefont{last\_name}

\begin{figure}[t]

\begin{lstlisting}
let wsp = [ \n\t\r]

@@

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_fields = perm (label, bib_author, bib_title) with (project (wsp+ . "," . wsp+) $\to$ ",\n")
let bibtex = "@book{" . bib_fields . "}"
\end{lstlisting}

\caption{QRE specification of \bibtex{} records. }
\label{fig:example-qre}
\end{figure}

The permutation QRE combinator, \kw{perm}, allows data to be unordered. For
example, the following instance of \kw{perm} allows the label, author and title
fields (which we assume have been defined earlier) to appear in any order among
the fields of a bibliography entry.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title)
\end{lstlisting}
%
To normalize any repeated separator between fields, one can extend a permutation
with any other QRE.  For instance, below, we normalize whitespace between fields,
leaving only comma followed by a single newline.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title) 
                 with (project (wsp+ . "," . wsp+) $\to$ ",\n")
\end{lstlisting}
%
The final QRE combinator is the \kw{normalize} combinator.  This combinator is an
``escape hatch'' that allows a programmer to manually define a function $f$
that sends each string matching a regular expression $R$ to some canonical
representative in another regular expression $R'$. The equivalence relation
defined by the normalize combinator is hence the equivalence relation defined by
the {\em fibres} of $f$; that is, for all strings $s$ and $s'$ that match $R$,
$s$ is equivalent to $s'$ if and only if $f(s) = f(s')$.

For instance, assume that $f(s)=$ `` '' for all whitespace strings $s$. Then the
project QRE \kw{p\_wsp} which matches any non-zero number of whitespace
characters and converts any such whitespace into a single space character could
be expressed using the \kw{normalize} combinator as
\begin{lstlisting}
let p_wsp = normalize (wsp+, " ", f)
\end{lstlisting}
However, although it is in some sense ``universal'' (see the discussion below
\saz{TODO}))the \kw{normalize} QRE imposes significant requirements on the
canonization function $f$, so it is usually simpler to use the other QRE
constructs where possible.

A QRE for the \bibtex{} format is presented in its
entirety in Figure~\ref{fig:example-qre}.

\subsection{QRE Lenses}
QREs allow users to specify regular languages modulo equivalence relations,
a first step in writing many data transformations.  However, writing the
inner bijective lens itself still takes a fair amount of work.  Each such lens
is constructed from simple primitive combinators including:
%
\begin{itemize}
  \item \kw{del} $s$: \textit{delete} the constant string $s$ in the get direction;
  insert it in the put direction.
  \item \kw{ins} $s$: \textit{insert} the constant string $s$ in the get direction;
  delete it in the put direction.
  \item \kw{copy} $R$: copy the text matching the regular expression $R$ in both
  directions.
\end{itemize}
In addition, such primitives may be combined using the regular operators
Kleene star, alternation and concatenation, or functional composition of
lenses.  Figure~\ref{fig:example-lens} illustrates the use of these
combinators to transform \bibtex{} to EndNote (and vice versa).
There is a fair amount of trickiness in writing these transformations, because
the programmer must keep mental track of what they have parsed so far \emph{in both
directions}, carefully inserting and deleting the relevant syntax on both
sides.

@@

%% The lens shown in Figure~\ref{fig:example-lens} transforms \bibtex{} data to
%% EndNote data  (and vice versa) by matching the delimiters and separators in a
%% bijective manner. For example, the \cd{preamble} lens replaces the \bibtex{}
%% preamble \verb|@Book| with the EndNote preamble \verb|%0 Book\n%F|. When the
%% programmer is writing the lens, she has to make sure that all of these small
%% deletions and insertions are correct, which can be a challenging task if the
%% lens is large enough, as in the definition of the lens \verb!author_lens! in
%% Figure~\ref{fig:example-lens}. This makes lens programming tedious and
%% error-prone.
To make this unsavory programming task more palatable, we identify a class of
quotient lenses that we call {\em QRE lenses} and show how to {\em synthesize}
them from a pair of QREs and a set of input-output examples. Our key
insight is to reduce the problem of synthesizing QRE lenses to the problem of
synthesizing {\em bijective lenses}~\cite{optician}, a simpler subset of
lenses that support strict bijections, rather than bijections modulo
equivalence relations.

@@

This technical investigation justifies replacing manually constructed lenses that may use
functional composition and nested canonizers with synthesized lenses in normal form.
The end result is that users may synthesize complex transformations with a single line
of code.  For instance, all the code in Figure~\ref{fig:example-lens} may be replaced by
a single call to the synthesis prodedure:
\begin{lstlisting}
let bib_to_end : (lens in bibtex $\Leftrightarrow$ endnote) =
    synth bibtex $\Leftrightarrow$ endnote  using  {(bib_example, end_example)}
\end{lstlisting}
\noindent Here, the generated  quotient lens synchronizes \cd{bibtex} and
\cd{endnote} formats.  It uses \cd{bib_example} and \cd{end_example} (the two
concrete example strings givn at the beginning of this section) to disambiguate.
In addition, as we saw earlier with the definition of \cd{s_to_c}, the synthesis
procedure itself can be used to create lenses that are in turn used to define
other QREs. The ability to mix QRE specification with QRE lens synthesis
yields a powerful and flexible way of creating bidirectional transformations.

\section{Quotient Regular Expressions}
\label{QRE}
\subsection{Syntax and Semantics of QREs}
Formally, the language of Quotient Regular Expressions (QREs) is given by the
following grammar:
\begin{align*}
c := \; &R \sep \project{R}{s} \sep \squash{c_1}{c_2}{f} \sep
\perm{c_1, \ldots, c_n}{c} \\
& | \; \normalize{R_1}{R_2}{f} \sep c_1 \semicolon c_2 \sep c_1 \cdot c_2 \sep (c_1 \sep
c_2) \sep c^*,
\end{align*}
where $c$ ranges over QREs, $R$ ranges over regular expressions, $f$ ranges over
functions between regular languages, and $s$ ranges over character strings.
